---
title: "Lab 4 MATH 342W"
author: "Mohammed Hasan"
output: pdf_document
date: "11:59PM February 29"
---


Create a dataset D which we call `Xy` such that the linear model has R^2 about 0\% but x, y are clearly associated.

```{r}
rm(list = ls())
x = seq(0, 6 * pi, length.out=1000) # 1000 output
y = sin(x)

pacman::p_load(ggplot2)

summary(lm(y ~ x))$r.squared

ggplot(data.frame(x = x, y = y)) + geom_point(aes(x = x, y = y))
```

Write a function `my_ols` that takes in `X`, a matrix with with p columns representing the feature measurements for each of the n units, a vector of n responses `y` and returns a list that contains the `b`, the p+1-sized column vector of OLS coefficients, `yhat` (the vector of n predictions), `e` (the vector of n residuals), `df` for degrees of freedom of the model, `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the R-squared metric). Internally, you cannot use `lm` or any other package; it must be done manually. You should throw errors if the inputs are non-numeric or not the same length. Or if `X` is not otherwise suitable. You should also name the class of the return value `my_ols` by using the `class` function as a setter. No need to create ROxygen documentation here.

df = degrees of freedom = p + 1 / number of dimensions / # of parameters


```{r}
# X -> Columns / features
my_ols = function(X, y){

  X = cbind(1, X)
  
  b = solve(t(X) %*% X) %*% t(X) %*% y

  y_hat = X %*% b
  
  e = y - y_hat

  SSE = sum(e^2)
  SST = sum((y - mean(y))^2)
  
  df = ncol(X)
  n = nrow(X)
  MSE = SSE / (n - df) 
  RMSE = sqrt(MSE)
  RSQ = (SST - SSE) / SST
  
  
  lmobj = list(
    b = b,
    y_hat = y_hat,
    e = e,
    SSE = SSE,
    SST = SST,
    df = df,
    MSE = MSE,
    RSQ = RSQ
  )
  
  class(lmobj) = "my_ols"
  lmobj
    
}
```

```{r}

```

Verify that the OLS coefficients for the `Type` of cars in the cars dataset gives you the same results as we did in class (i.e. the ybar's within group). 

```{r}
#TO-DO
cars = MASS::Cars93
cars_X = model.matrix(~Type, cars)
head(cars_X)

cars_y = cars$Price 

my_ols(cars_X[, -1], cars_y)
cat("...")
```


Create a prediction method `g` that takes in a vector `x_star` and the dataset D i.e. `X` and `y` and returns the OLS predictions. Let `X` be a matrix with with p columns representing the feature measurements for each of the n units

```{r}
g = function(x_star, X, y){
  c(1,x_star) %*% my_ols(X,y)$b  
}
```


Load up the famous iris dataset. We are going to do a different prediction problem. Imagine the only input x is Species and you are trying to predict y which is Petal.Length. A reasonable prediction is the average petal length within each Species. Prove that this is the OLS model by fitting an appropriate `lm` and then using the predict function to verify.

```{r}
data(iris)
coef(lm(Petal.Length ~ Species, iris))

mean(iris$Petal.Length[iris$Species == "setosa"]) 
mean(iris$Petal.Length[iris$Species == "versicolor"]) 
mean(iris$Petal.Length[iris$Species == "virginica"])

```

Construct the design matrix with an intercept, X without using `model.matrix`.

```{r}
# # design matrix == x matrix
cbind(1, ifelse(iris$Species == "versicolor", 1, 0), ifelse(iris$Species == "virginica", 1, 0))

```

We now load the diamonds dataset. Skim the dataset using skimr or summary. What is the datatype of the color feature? : ORDERED FACTOR 


```{r}
rm(list = ls())
pacman::p_load(ggplot2, skim)
pacman::p_load(skim)
diamonds = ggplot2::diamonds

summary(diamonds)

typeof(diamonds$color)

```

Find the levels of the color feature.

```{r}
levels(diamonds$color)
```

Create new feature in the diamonds dataset, `color_as_numeric`, which is color expressed as a continuous interval value. 

```{r}
# NUMERIC
diamonds$color_as_numeric = as.numeric(diamonds$color)
head(diamonds)

# NOMINAL 
diamonds$color_as_nominal = factor(diamonds$color)
head(diamonds)
```

Use that converted feature as the one predictor in a regression. How well does this regression do as measured by RMSE?

```{r}
#TO-DO
# Trying to model price using color as numeric
diamonds_coeff = lm(price ~ color_as_numeric, diamonds) # gets w_0, w_1
summary (diamonds_coeff)$sigma # RSQ
```


Create new feature in the diamonds dataset, `color_as_nominal`, which is color expressed as a nominal categorical variable. 

```{r}
#TO-DO
diamonds_coeff_nominal = lm(price ~ color_as_nominal, diamonds)
summary (diamonds_coeff_nominal)$sigma
```

Use that converted feature as the one predictor in a regression. How well does this regression do as measured by RMSE?

```{r}
summary(lm(price~color_as_nominal, diamonds))$sigma
```

Which regression does better - `color_as_numeric` or `color_as_nominal`? Why?

color_as_numeric because it is more accurate

Return to the iris dataset. Find the hat matrix H for this regression of diamond price on diamond color. Use only the first 1,00 observations in the diamond dataset.

```{r}
rm(list = ls())
diamonds1000 = ggplot2::diamonds[1:1000,]
# WE NEED TO FIND THE X MATRIX => H = X(X^T X)^-1 X^T
X_diamonds = model.matrix(price ~ color, diamonds1000) #Regress price to color
H = X_diamonds %*% solve(t(X_diamonds) %*% X_diamonds) %*% t(X_diamonds)
```

Verify this hat matrix is symmetric using the `expect_equal` function in the package `testthat`.

```{r}
pacman::p_load(testthat)
expect_equal(H, t(H)) 
```

Verify this hat matrix is idempotent using the `expect_equal` function in the package `testthat`.

```{r}
pacman::p_load(testthat)
#TO-DO
expect_equal(H %*% H, H)
```

Using the `diag` function, find the trace of the hat matrix.

```{r}
#TO-DO
sum(diag(H))
# Trace of an orthogonal matrix is its rank 
```

It turns out the trace of a hat matrix is the same as its rank! But we don't have time to prove these interesting and useful facts..


Using the hat matrix, compute the yhat vector and using the projection onto the residual space, compute the e vector and verify they are orthogonal to each other.

```{r}
#TO-DO

y_diamond = diamonds1000$price
yhat_diamond = H%*%y_diamond
yhat_diamond

I = diag(nrow(H))
e = (I - H) %*% y_diamond


t(e) %*% yhat_diamond  # 0 -> but since the inversion the bits are off and giving us an error

```

Compute SST, SSR and SSE and R^2 and then show that SST = SSR + SSE.

```{r}

SST = sum((y_diamond - mean(y_diamond))^2)
SSE = sum(e^2)
SSR = sum((yhat_diamond - mean(y_diamond))^2)

SST - sum(SSR + SSE)

RSQ = SSR / SST
RSQ

```

Find the angle theta between y - ybar 1 and yhat - ybar 1 and then verify that its cosine squared is the same as the R^2 from the previous problem.

```{r}
#TO-DO

numer = sqrt(sum((yhat_diamond - mean(y_diamond))^2))
denom = sqrt(sum((y_diamond - mean(y_diamond))^2))
theta = acos(numer/denom)
cos(theta)^2
```

Project the y vector onto each column of the X matrix and test if the sum of these projections is the same as yhat.

```{r}
n = nrow(X)
sum_proj_y = matrix(0, nrow = n, ncol = 1)
for(
  j in 1:ncol(X)
){
  x_j = X[,j, drop = FALSE]
  sum_proj_y = sum_proj_y + x_j %% t(x_j) / sum(x_j^2) %% y
}
expect_equal(yhat, sum_proj_y)
```

Convert this design matrix into Q, an orthonormal matrix.

```{r}
Q = matrix(NA, nrow = nrow(X), ncol = ncol(X))
Q[,j] = X[,1]
for (j in 2:ncol(X)) {
  Q[,j] = X[,j]
  for (k in 1: (j-1)) {
    q_k = Q[,k, drop = FALSE]
    Q[,j] = Q[,j] - (q_k %% t(q_k) / sum(q_k^2)) %% X[,j]
  }
}
for (j in 1: ncol(X)){
  Q[,j] = Q[,j] / sqrt(sum(Q[,j]^2))
}
```

Project the y vector onto each column of the Q matrix and test if the sum of these projections is the same as yhat.

```{r}
n = nrow(X)
sum_proj_y = matrix(0, nrow = n, ncol = 1)
for(
  j in 1:ncol(X)
){
  x_j = X[,j, drop = FALSE]
  sum_proj_y = sum_proj_y + x_j %% t(x_j) / sum(x_j^2) %% y
}
expect_equal(yhat, sum_proj_y)
```

Find the p=3 linear OLS estimates if Q is used as the design matrix using the `lm` method. Is the OLS solution the same as the OLS solution for X?

```{r}
n = nrow(X)
sum_proj_y = matrix(0, nrow = n, ncol = 1)
for(
  j in 1:ncol(X)
){
  q_j = Q[,j, drop = FALSE]
  sum_proj_y = sum_proj_y + q_j %% t(q_j) / sum(q_j^2) %% y
}
```

Use the predict function and ensure that the predicted values are the same for both linear models: the one created with X  as its design matrix and the one created with Q as its design matrix.

```{r}
mod_vanilla = coef(lm(y ~0+ X))
b = coef(mod_vanilla)
b
mod_ortho = coef(lm(y ~0+ Q))
b_q = coef(mod_ortho)
b_q
```


Clear the workspace and load the boston housing data and extract X and y. The dimensions are n = 506 and p = 13. Create a matrix that is (p + 1) x (p + 1) full of NA's. Label the columns the same columns as X. Do not label the rows. For the first row, find the OLS estimate of the y regressed on the first column only and put that in the first entry. For the second row, find the OLS estimates of the y regressed on the first and second columns of X only and put them in the first and second entries. For the third row, find the OLS estimates of the y regressed on the first, second and third columns of X only and put them in the first, second and third entries, etc. For the last row, fill it with the full OLS estimates.

```{r}
pacman::p_load(MASS)

X <- as.matrix(Boston[, -ncol(Boston)]) # All columns except the target variable
y <- Boston$medv

# Initialize parameters
n <- nrow(X)
p <- ncol(X)

# Create a matrix of NA's
result_matrix <- matrix(NA, nrow = p + 1, ncol = p + 1)
colnames(result_matrix) <- c(colnames(X), "Intercept")

# Loop through each subset of columns and perform OLS regression, filling the matrix
for (i in 1:p) {
  X_subset <- X[, 1:i, drop = FALSE] # Ensure it remains a matrix
  model <- lm(y ~ ., data = as.data.frame(X_subset))
  
  # Fill the matrix with coefficients; for the first model, include the intercept
  if (i == 1) {
    result_matrix[1, 1] <- coef(model)[1] # Assuming you want the intercept in the first cell
  }
  
  # Adjusting for intercept handling
  # Intercept is placed in the last column for each model
  result_matrix[i + 1, 1:i] <- coef(model)[-1] # Coefficients
  result_matrix[i, p + 1] <- coef(model)[1] # Intercept at the end for each row
}

# For the full model
full_model <- lm(y ~ ., data = Boston)
result_matrix[p + 1, 1:p] <- coef(full_model)[-1] # Assign coefficients except intercept
result_matrix[p + 1, p + 1] <- coef(full_model)[1] # Intercept in the last position

# Print the result matrix
print(result_matrix)
```

Why are the estimates changing from row to row as you add in more predictors?

#TO-DO

Create a vector of length p+1 and compute the R^2 values for each of the above models. 

```{r}
model_numeric = lm(price ~ color_as_numeric, data = diamonds)
model_nominal = lm(price ~ color_as_nominal, data = diamonds)

r_squared_numeric = summary(model_numeric)$r.squared
r_squared_nominal = summary(model_nominal)$r.squared

r_squared_vector = c(r_squared_numeric, r_squared_nominal)

p = length(levels(diamonds$color))
r_squared_vector_full = c(r_squared_vector, rep(NA, p - 1))

r_squared_vector_full
```

Is R^2 monotonically increasing? Why?

#TO-DO

Create a 2x2 matrix with the first column 1's and the next column iid normals. Find the absolute value of the angle (in degrees, not radians) between the two columns in absolute difference from 90 degrees.

```{r}
n = 100

X = matrix(rnorm(2 * n), ncol = 2)
acos(t(X[,1]) %*% X[,2] / sqrt(sum(X[, 1]^2) * sum(X[, 2]^2))) * 180 / pi
```

Repeat this exercise `Nsim = 1e5` times and report the average absolute angle.

```{r}
set.seed(123)
Nsim = 1e5
n = 100
angles = numeric(Nsim)

for(i in 1:Nsim) {

  X = matrix(nrow = n, ncol = 2)
  X[,1] = rep(1, n)
  X[,2] = rnorm(n)

  angle == acos(sum(X[,1] * X[,2]) / (sqrt(sum(X[,1]^2)) * sqrt(sum(X[,2]^2)))) * (180 / pi)

  angles[i] = abs(90 - angle)
}

average_absolute_angle = mean(angles)
average_absolute_angle
```

Create a n x 2 matrix with the first column 1's and the next column iid normals. Find the absolute value of the angle (in degrees, not radians) between the two columns. For n = 10, 50, 100, 200, 500, 1000, report the average absolute angle over `Nsim = 1e5` simulations.

```{r}
calculate_angle_difference = function(n, Nsim) {
  angles = numeric(Nsim)

  for (i in 1:Nsim) {
    X = matrix(nrow = n, ncol = 2)
    X[, 1] = rep(1, n)
    X[, 2] = rnorm(n)

    angle = acos(sum(X[, 1] * X[, 2]) / (sqrt(sum(X[, 1]^2)) * sqrt(sum(X[, 2]^2)))) * (180 / pi)

    angles[i] = abs(90 - angle)
  }

  mean(angles)
}

n_values = c(10, 50, 100, 200, 500, 1000)
Nsim = 1e5

average_angles = numeric(length(n_values))

for (i in seq_along(n_values)) {
  average_angles[i] = calculate_angle_difference(n_values[i], Nsim)
}

results = data.frame(n = n_values, Average_Absolute_Angle = average_angles)

print(results)
```

What is this absolute angle difference from 90 degrees converging to? Why does this make sense?

This convergence reflects the increasing orthogonality between a fixed direction and a direction determined by drawing from a high-dimensional Gaussian distribution resulting it to become orthogonal to any fixed direction as the dimensionality increases.

