---
title: "Lab 7"
author: "Mohammed Zohair Hasan"
output: pdf_document
---


#Polynomial Regression and Interaction Regression

We will work with the diamonds dataset again. Here we load up the dataset and convert all factors to nominal type:

```{r}
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)      #convert to nominal
diamonds$color =    factor(diamonds$color, ordered = FALSE)    #convert to nominal
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)  #convert to nominal
skimr::skim(diamonds)
```

Given the information above, what are the number of columns in the raw X matrix?

There are 9 columns

Verify this using code:

```{r}
X = diamonds[, !names(diamonds) %in% c("price")]
print(ncol(X))
```

Would it make sense to use polynomial expansions for the variables cut, color and clarity? Why or why not?

no; categorical 

Would it make sense to use log transformations for the variables cut, color and clarity? Why or why not?

no; 

In order to ensure there is no time trend in the data, randomize the order of the diamond observations in D:.

```{r}
diamonds = diamonds[sample(1:nrow(diamonds)), ]
```

Let's also concentrate only on diamonds with <= 2 carats to avoid the issue we saw with the maximum. So subset the dataset. Create a variable n equal to the number of remaining rows as this will be useful for later. Then plot it.

```{r}
diamonds = diamonds[diamonds$carat<=2, ]
n = nrow(diamonds)
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```

Create a linear model of price ~ carat and gauge its in-sample performance using s_e.

```{r}
summary(lm(price ~ carat, data = diamonds))$sigma
```

Create a model of price ~ clarity and gauge its in-sample performance

```{r}
summary(lm(price ~ clarity, data = diamonds))$sigma
```

Why is the model price ~ carat substantially more accurate than price ~ clarity?

The difference in data type; clarity is a categorical variable while carat is continuous.

Create a new transformed feature ln_carat and plot it vs price.

```{r}
diamonds$ln_carat = log(diamonds$carat)
ggplot(diamonds, aes(x = ln_carat, y = price)) + 
  geom_point()
```

Would price ~ ln_carat be a better fitting model than price ~ carat? Why or why not?

No, there is less of a linear pattern.

Verify this by comparing R^2 and RMSE of the two models:

```{r}
mod1=lm(price ~ carat, data = diamonds)
mod2=lm(price ~ ln_carat, data = diamonds)
summary(mod1)$r.squared
summary(mod1)$sigma
summary(mod2)$r.squared
summary(mod2)$sigma
```

Create a new transformed feature ln_price and plot its estimated density:


```{r}
diamonds$ln_price = log(diamonds$price)
ggplot(diamonds) + geom_histogram(aes(x = ln_price), binwidth = 0.01)
```


Now plot it vs carat.

```{r}
ggplot(diamonds, aes(x = carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ carat be a better fitting model than price ~ carat? Why or why not?

Yes! It looks more linear.

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
mod1=lm(price ~ carat, data = diamonds)
mod2=lm(ln_price ~ carat, data = diamonds)
summary(mod1)$r.squared
summary(mod1)$sigma
y_hat = exp(mod2$fitted.values)
y = diamonds$price
e = y - y_hat
1 - sum(e^2)/sum((y - mean(y))^2)
sqrt(sum(e^2)/(mod2$df.residual))
```

We just compared in-sample statistics to draw a conclusion on which model has better performance. But in-sample statistics can lie! Why is what we did valid?

There is a high n and a low p so the in-sample estimation will be reliable. 

Plot ln_price vs ln_carat.

```{r}
ggplot(diamonds, aes(x = ln_carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ ln_carat be the best fitting model than the previous three we considered? Why or why not?

Yes, because it seems to be the most linear model that we can get with using logs.

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
mod2=lm(ln_price ~ ln_carat, data = diamonds)
y_hat = exp(mod2$fitted.values)
y = diamonds$price
e = y - y_hat
1 - sum(e^2)/sum((y - mean(y))^2)
sqrt(sum(e^2)/(mod2$df.residual))
```

Compute b, the OLS slope coefficients for this new model of ln_price ~ ln_carat.

```{r}
coef(mod2)
```

Interpret b_1, the estimated slope of ln_carat.

if carat increases by p-percent, then price increases by p-percent*1.7

Interpret b_0, the estimated intercept.

for a 1-carat diamond, the price is predicted to be about e^8.5

Create other features ln_x, ln_y, ln_z, ln_depth, ln_table.

```{r}
diamonds$ln_x = log(diamonds$x)
diamonds$ln_y = log(diamonds$y)
diamonds$ln_z = log(diamonds$z)
diamonds$ln_depth = log(diamonds$depth)
diamonds$ln_table = log(diamonds$table)
```

From now on, we will be modeling ln_price (not raw price) as the prediction target. 

Create a model (B) of ln_price on ln_carat interacted with clarity and compare its performance with the model (A) ln_price ~ ln_carat.

```{r}
#Model B
mod_b = lm(ln_price ~ ln_carat*clarity, diamonds)
summary(mod_b)$sigma
```

Which model does better? Why?

Model B is better. It has more features so we reduced ignorance error.

Create a model of (C) ln_price on ln_carat interacted with every categorical feature (clarity, cut and color) and compare its performance with model (B)

```{r}
#Model C
mod_c = lm(ln_price ~ ln_carat*(clarity + cut + color), diamonds)
summary(mod_c)$sigma
```

Which model does better? Why?

Model C because it captures more variance in the data.

Create a model (D) of ln_price on every continuous feature (logs of carat, x, y, z, depth, table) interacted with every categorical feature (clarity, cut and color) and compare its performance with model (C).

```{r}
#Model D
diamonds = diamonds[diamonds$x != 0 & diamonds$y != 0 & diamonds$z != 0 & diamonds$depth != 0 & diamonds$table != 0, ]
mod_d = lm(ln_price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table)*(clarity + cut + color), diamonds)
summary(mod_d)$sigma
```

Which model does better? Why?

Depending the complexity we introduced in model D, if this helps see more variance. We can conclude model D works better.

What is the p of this model D? Compute with code.

```{r}
mod_d$rank - 1
```

Create model (E) which is the same as before except create include the raw features interacted with the categorical features and gauge the performance against (D).

```{r}
#Model E
mod_e = lm(ln_price ~ (carat + x + y + z + depth + table)*(clarity + cut + color), diamonds)
summary(mod_e)$sigma
```

Which model does better? Why?

Model E

Create model (F) which is the same as before except also include also third degree polynomials of the continuous features interacted with the categorical features and gauge performance against (E). By this time you're getting good with R's formula syntax!

```{r}
#Model F
mod_f = lm(ln_price ~ (poly(carat, 3) + poly(x, 3) + poly(y, 3) + poly(z, 3) + poly(depth, 3) + poly(table, 3))*(clarity + cut + color), diamonds)
summary(mod_f)$sigma
```

Which model does better? Why?

Model F because it includes third degree polynomials of the continuous features

We can add more features

We should probably assess oos performance now. Sample 2,000 diamonds and use these to create a training set of 1,800 random diamonds and a test set of 200 random diamonds. Define K and do this splitting:

```{r}
train_indices = sample(1:nrow(diamonds), 1800)
test_indices = sample(setdiff(1:nrow(diamonds), train_indices))
diamonds_train = diamonds[train_indices,]
diamonds_test = diamonds[test_indices,]

mod_f = lm(ln_price ~ (poly(carat, 3) + poly(x, 3) + poly(y, 3) + poly(z, 3) + poly(depth, 3) + poly(table, 3))*(clarity + cut + color), diamonds_train)

y_hat_test = predict(mod_f, diamonds_test)
mean((diamonds_test$ln_price - y_hat_test)^2)
```

Compute in and out of sample performance for models A-F. Use s_e as the metric (standard error of the residuals). Create a list with keys A, B, ..., F to store these metrics. Remember the performances here will be worse than before since before you're using nearly 52,000 diamonds to build a model and now it's only 1,800! 

```{r}
performance_metrics = list()

mod_a <- lm(ln_price ~ ln_carat, data = diamonds_train)
pred_a_train = predict(mod_a, newdata = diamonds_train)
pred_a_test = predict(mod_a, newdata = diamonds_test)
mse_a_train = mean((diamonds_train$ln_price - pred_a_train)^2)
mse_a_test = mean((diamonds_test$ln_price - pred_a_test)^2)
s_e_a_train = sqrt(mse_a_train)
s_e_a_test = sqrt(mse_a_test)
performance_metrics[["A"]] = list(in_sample = s_e_a_train, out_of_sample = s_e_a_test)

```

You computed oos metrics only on n_* = 200 diamonds. What problem(s) do you expect in these oos metrics?

High variability due to only 200 diamonds.

To do the K-fold cross validation we need to get the splits right and crossing is hard. We've developed code for this already in a previous lab.

```{r}
temp = rnorm(n)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
rm(temp)
head(folds_vec, 200)
```

Do the K-fold cross validation for model F and compute the overall s_e and s_s_e. 

```{r}
K = 10 # amt of folds for k
n = nrow(diamonds)

set.seed(123)
temp <- rnorm(n)
folds_vec <- cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
rm(temp)

total_sse = 0
total_n = 0

for (k in 1:K) {
  test_indices <- which(folds_vec == k)
  train_indices <- setdiff(1:n, test_indices)
  
  diamonds_train <- diamonds[train_indices, ]
  diamonds_test <- diamonds[test_indices, ]
  
  mod_f <- lm(ln_price ~ (poly(carat, 3) + poly(x, 3) + poly(y, 3) + poly(z, 3) + poly(depth, 3) + poly(table, 3)) * (clarity + cut + color), data = diamonds_train)
  
  predictions <- predict(mod_f, newdata = diamonds_test)
  
  sse <- sum((diamonds_test$ln_price - predictions)^2)
  total_sse <- total_sse + sse
  total_n <- total_n + length(test_indices)
}

s_s_e <- total_sse
s_e <- sqrt(total_sse / total_n)

cat("Overall s_s_e:", s_s_e, "\n")
cat("Overall s_e:", s_e, "\n")
```

Does K-fold CV help reduce variance in the oos s_e? Discuss.

Yes, the point of it is to give a better idea of how the model will perform on unseen data.

Imagine using the entire rest of the dataset besides the 2,000 training observations divvied up into slices of 200. Measure the oos error for each slice on Model F in a vector `s_e_s_F` and compute the `s_s_e_F` and also plot it.

```{r}
ggplot(data.frame(s_e_s_F = s_e_s_F)) + geom_histogram(aes(x = s_e_s_F))
```


#Rcpp and optimizing R

Write a function `dot_product_R` in R that takes in two vectors `v1` and `v2` and returns their dot product.

```{r}
dot_product_R = function(v1, v2){
  dot_product = 0 
  for(i in 1:length(v1)){
    dot_product = dot_product + v1[i] * v2[i] 
    
  }
  dot_product
}
```

Write a function `dot_product_cpp` in C++ and make sure it compiles.
```{r}
library(Rcpp)
cppFunction('int add(int x, int y) { return x + y; }')

```

```{r}
pacman::p_load(Rcpp)


cppFunction('
  double dot_product_cpp(NumericVector v1, NumericVector v2) {
  
    double dot_product = 0; 
    for(int i = 0; i < v1.length(); i++) {
      dot_product = dot_product + v1[i] * v2[i]; 
    }
   return dot_product;
  }
')
```

Create two vectors of standard normal realizations with length `n=1e6` and test the different in speed.

```{r}
n = 1e6           
v1 = rnorm(n)
v2 = rnorm(n)

pacman::p_load(microbenchmark)
microbenchmark(
  dot_product_R(v1, v2), 
  dot_product_cpp(v1, v2),
  times = 10
)
```

Implement the Gram Schmidt routine as a C++ function `gram_schmidt_cpp`.

```{r}
library(Rcpp)

cppFunction('
NumericMatrix gram_schmidt_cpp(NumericMatrix V) {
  int n = V.nrow(), k = V.ncol();
  NumericMatrix U(n, k);
  
  for(int i = 0; i < k; ++i) {
    NumericVector v = V(_, i);
    U(_, i) = v;
    for(int j = 0; j < i; ++j) {
      NumericVector u_j = U(_, j);
      // Projecting v onto u_j
      double dot = std::inner_product(v.begin(), v.end(), u_j.begin(), 0.0);
      double norm_u_j = std::inner_product(u_j.begin(), u_j.end(), u_j.begin(), 0.0);
      // Subtract the projection from the current vector
      U(_, i) = U(_, i) - (dot / norm_u_j) * u_j;
    }
    NumericVector u_i = U(_, i);
    double norm_u_i = sqrt(std::inner_product(u_i.begin(), u_i.end(), u_i.begin(), 0.0));
    U(_, i) = U(_, i) / norm_u_i;
  }
  return U;
}')
```

Here is the implementation in R for reference taken from lab 5:

```{r}
gram_schmidt_R = function(X){
  #first create orthogonal matrix
  V = matrix(NA, nrow = nrow(X), ncol = ncol(X))
  V[, 1] = X[, 1]
  
  for (j in 2 : ncol(X)){
    V[, j] = X[, j]
    
    for (k in 1 : (j-1)){
      v_k = V[, k, drop = FALSE]
      V[, j] = V[, j, drop = FALSE] - (t(t(v_k)) %*% t(v_k) / sum(v_k^2)) %*% t(t(X[, j])) #i.e. the orthogonal projection of X[, j] onto v_k
    }
  }
  
  Q = matrix(NA, nrow = nrow(X), ncol = ncol(X))
  for (j in 1 : ncol(X)){
    Q[, j] = V[, j] / sqrt(sum(V[, j]^2))
  }
  Q
}
```

Now let's see how much faster C++ is by running it on the boston housing data design matrix
```{r}
X = model.matrix(medv ~ ., MASS::Boston)

microbenchmark(
  gram_schmidt_R(X),
  gram_schmidt_cpp(X),
  times = 10
)
```


